\chapter*{Preface}
\addcontentsline{toc}{chapter}{Preface}


\noindent\textbf{TO DO:} Add full persistence with node splitting to
the chapter on persistence.


This document is intended to act as course notes for the course that
currently goes by the name \emph{COMP5408: Advanced Data Structures}.
The course is about techniques for the design and analysis of data
structures.  The focus of the course is on general techniques that are
simple, easy to understand, implementable and that lead to data
structures with good worst-case or expected running times.

The intended audience for this course is anyone who has taken an
undergraduate course in algorithms and data structures.  I expect that
students understand big-$O$ notation and have some experience with its
use in analyzing algorithms.  It would also be helpful to have some
experience with basic probability and the idea of \emph{amortized
analysis}, but it is not strictly necessary.

The bulk of this text is about dictionaries and priority queues.  For
those who would like to use this text as a reference, rather than read
it cover-to-cover, Tables~\ref{tab:dictionaries} and \ref{tab:pq}
summarize the dictionary and priority queue data structures that are
covered in these notes.  Dictionaries are classified primarily based
on whether they are comparison-based or integer-based and secondarily
on whether they are meant to store fixed-length or variable-length
keys.  Priority queues are characterized based on the running times of
various priority queue operations.  The names of the data structures are
suffixed with an (E), (A) or (W) depending on whether they are efficient
in an expected, amortized, or worst-case sense, respectively.

\begin{table}
\begin{center}
\begin{tabular}{|l|l|l|l|l|l|}\hline
Comparisons & Key size & Data Structure & Chapter \\ \hline\hline
Comparison-based & Fixed  & Treaps (E) & \ref{chap:records} \\
                 &                 & Skiplists (E) & \ref{chap:fractional} \\
                 &                 & Scapegoat trees (A) & \ref{chap:laziness} \\
                 &                 & Working set (W) & \ref{chap:entropy} \\
                 &                 & Splay trees (A) & \ref{chap:adjusting} \\
                 & Variable  & Ternary tries (E/W) & 
			\ref{chap:strings} \\ \hline
Integer-based    & Fixed  & van Emde Boas trees (W) & 
			\ref{chap:bounded} \\
                 &                 & Willard's trees (W) & 
			\ref{chap:bounded} \\
                 & Variable  & Tries (W) & \ref{chap:strings} \\
                 &                    & PATRICIA trees (W) & \ref{chap:strings}
			\\ \hline 
\end{tabular}
\end{center}
\caption{Summary of dictionaries.}\tablabel{dictionaries}
\end{table}

\begin{table}
\begin{center}
\begin{tabular}{|l|l|l|l|l|l|l|l|}\hline
Data structure & Chp. & FindMin & DeleteMin & Insert & Meld & Delete & Decrease-Key \\ \hline\hline
Binary heaps (W) &  & $1$ & $\log n$ & $\log n$ & X & $\log n$ & $\log n$ \\
Heaters (E) & \ref{chap:records} & $1$ & $\log n$ & $\log n$ & X & $\log n$ & $\log n$ \\
Pairing-heaps (A) & \ref{chap:adjusting} & $1$ & $\log n$ & $1$ & $1$ & $\log n$ & $\log n$ \\
Queaps (A) & \ref{chap:adjusting} & $1$ & $\log n$ & $1$ & X & $\log n$ & $\log n$ \\
Leftist heaps (W) & \ref{chap:imbalance} & $1$ & $\log n$ & $\log n$ & $\log n$ & X & X \\
Randomized heaps (E) & \ref{chap:imbalance} & $1$ & $\log n$ & $\log n$ & $\log n$ & X & X \\
Skew heaps (A) & \ref{chap:imbalance} & $1$ & $\log n$ & $\log n$ & $\log n$ & $\log n$ & $\log n$ \\
Binary-counting heaps (A) & \ref{chap:binary} & $1$ & $\log n$ & $\log n$ & $\log n$ & $\log n$ & $\log n$ \\
Binomial heaps (W) & \ref{chap:binary} & $1$ & $\log n$ & $\log n$ & $\log n$ & $\log n$ & $\log n$ \\
Fibonacci heaps (A) & \ref{chap:binary}  & $1$ & $\log n$ & $1$ & $1$  & $\log n$ & $1$ \\
Min-max heaps (W) & \ref{chap:more-heaps}  & $1$ & $\log n$ & $\log n$ & X & $\log n$ & $\log n$ \\
H.-O. stacks (W) & \ref{chap:more-heaps}  & $1$ & X & X & X  & X & X \\
H.-O. queues (W) & \ref{chap:more-heaps}  & $1$ & X & X & X  & X & X \\
H.-O. deques (W) & \ref{chap:more-heaps}  & $1$ & X & X & X  & X & X \\ \hline
\end{tabular}
\end{center}
\caption{Summary of priority queues.}\tablabel{pq}
\end{table}




These notes are still a work in progress. Many of the chapters are
unfinished (even unstarted) and probably contain many errors.  I
intend to continue finishing chapters and correcting errors as the
course progresses.  I would also like to get as much feedback as
possible about any errors, omissions or inconsistencies that readers
might find.  Anyone offering helpful comments will have their name
appear in the acknowledgements section of future editions of the
notes.  For now, I would like to thank the students of 95.590X W2002:

\mbox{Ghada Badr}, \mbox{Paul Boone}, \mbox{Steven Bowles},
\mbox{Xiaoli Chen}, \mbox{Ralf Dagher}, \mbox{Andras Frankel},
\mbox{Bo Han}, \mbox{Jon Harris}, \mbox{Xiaofei Jia}, \mbox{Xiaobing
Jin}, \mbox{Aaron Lee}, \mbox{Rocky Ng}, \mbox{Sylvain Pitre},
\mbox{Cynthia Sun}, \mbox{Haibo Wang}, and \mbox{Juemin Yu}.

